\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Local Adaptive Optimization of Time Step}

### 2013-10-31 report/article.tex
Report figures generated by runReport and report section experiments for switch and atomic (half done)

We have now fully described the simple case of simulating a mass-spring system
using the explicit Euler method and the next subsections describe the three
different methods/modifications for applying LAOTS to the simulation just
described.

\subsection{Adapt-on-switch}
The first LAOT method we propose is time-adaptive. We perform a simulation
with an initial timestep size. In each step we try to predict whether a timestep
of the same size using the current velocities will result in a switch of the
left and right vertices of any of the springs. If this is the case, we should
try to avoid the switch by decreasing the timestep of the current iteration,
re-calculate the velocity and position changes and resume stepping of the
simulation using the new timestep size. When switches aren't detected anymore
we gradually increase the timestep of the simulation again. We call this
method the adapt-on-switch method.

@@

\subsection{Atomic updating using heuristics}
The second LAOT method we propose is space-adaptive. We perform simulations
using a fixed timestep size. In each iteration we perform position update of
all vertices, but only re-calculate acceleration and velocity of the springs
and points that fulfill some defined heuristics.

The heuristics can be chosen to fit the domain of the simulation, but we have
chosen to use the number of steps since last update and the ratio between the
current spring length and the spring length at the last update. This should
force a bound on the maximal timestep before update as well as making sure
updates are done more frequently in case the springlength is changing more
rapidly. Further details are explained in section \ref{sec:experiments}.

@@

\todo{Write something about the explicit euler based on uniform}
\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/standard_uniform_30fps.png}
        \caption{30 fps}
        \label{fig:std_uni_30fps}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/standard_uniform_1fps.png}
        \caption{1 fps}
        \label{fig:std_uni_1fps}
    \end{subfigure}
    \caption{Standard simulations of particles distributed uniformly using
    different timestep sizes}
\end{figure}

\subsection{Adapt-on-switch}
Special things to note: How subdivision is done: new fixed (30) or relative (dt / 10)
new fixed: We don't catch it early enough. Maybe something with the explicit
euler, maybe something with the linear model. We should catch the error before
it escalates into a state in which we cannot rescue it anymore.

Relative: We get too fine-grained - maybe related to the above
\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/switch_multiscale_1fps_relative.png}
        \caption{Relative timestepping ($\Delta t = \frac{\Delta t}{100}$)}
        \label{fig:switch_multi_1fps_relative}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/switch_multiscale_1fps_absolute.png}
        \caption{Absolute timestepping ($\Delta t = \frac{1}{30}$)}
        \label{fig:switch_multi_1fps_absolute}
    \end{subfigure}
    \caption{Adapt-on-switch simulation of particles distributed with two
    scales using 1 fps}
\end{figure}

\subsection{Atomic updating using heuristics}

Heuristic: steps since last update and ratio of length since last update

interpolation of positions makes the integration more stable since the stepping
is based on a point closer in time

@@

The heuristic dampens the system

>>>>

We have now fully described the simple case of simulating a mass-spring system
using the explicit Euler method and the next subsections describe the three
different methods/modifications for applying LAOTS to the simulation just
described. The three methods are using different ideas to implement LOATS. They
could however possibly be combined in case they turn out to be working as
intended to achieve the best possible LAOTS scheme.

\subsection{Adapt-on-switch}
\label{sec:method_switch}
The first LAOT method we propose is time-adaptive. We perform a simulation
with an initial timestep size. In each step we try to predict whether a timestep
of the same size using the current velocities will result in a switch of the
left and right vertices of any of the springs. If this is the case, we should
try to avoid the switch by decreasing the timestep of the current iteration,
re-calculate the velocity and position changes and resume stepping of the
simulation using the new timestep size. When switches aren't detected anymore
we gradually increase the timestep of the simulation again. We call this
method the adapt-on-switch method.

@@

\subsection{Atomic Updating using Heuristics}
The second LAOT method we propose is space-adaptive. We perform simulations
using a fixed timestep size. In each iteration we perform position update of
all vertices, but only re-calculate acceleration and velocity of the springs
and points that fulfill some defined heuristics and hence the method is called
``Atomic Updating using Heuristics'' (AUH).

The heuristics can be chosen to fit the domain of the simulation, but we
have chosen to use the number of steps since last update, $n_{last}$, and
the ratio between the current spring length and the spring length at the
last update $r_{last}$. This should force a bound on the maximal timestep
before update as well as making sure updates are done more frequently in case
the springlength is changing more rapidly. Further details are explained in
section \ref{sec:experiments}.

@@

\todo{Write something about the explicit euler based on uniform example}
%\begin{figure}[H]
%    \begin{subfigure}{0.5\textwidth}
%        \includegraphics[width=\textwidth]{../images/standard_uniform_30fps.png}
%        \caption{30 fps}
%        \label{fig:std_uni_30fps}
%    \end{subfigure}
%    \begin{subfigure}{0.5\textwidth}
%        \includegraphics[width=\textwidth]{../images/standard_uniform_1fps.png}
%        \caption{1 fps}
%        \label{fig:std_uni_1fps}
%    \end{subfigure}
%    \caption{Standard simulations of particles distributed uniformly using
%    different timestep sizes}
%\end{figure}

\subsection{Adapt-on-switch}
As mentioned in section \ref{sec:method_switch} the adapt-on-switch method
can be implemented using either a relative or a absolute adaption of the
timestep. In our experiments we have tested out both methods defining the
relative adaption to scale the timestep by a factor of $\frac{1}{100}$ and
the absolute solution to set the timestep to $\frac{1}{30}$ when switches are
detected since this generated nice simulations for the standard simulation .
Furthermore the up-scaling for the relative scheme is set to a factor of 10
since we wish to avoid running into new switches immediately.

Figure \ref{fig:switch} shows the result of running simulations
described earlier at 1 fps using the adapt-on-switch scheme, where
the red crosses mark timesteps where a switch is predicted in the
following step and hence the schemes are ``activated''. Figure
\ref{fig:switch_multi_1fps_relative} is simulated using the relative scheme
and figure \ref{fig:switch_multi_1fps_absolute} is simulated using the
absolute scheme. As we clearly see in both figures the schemes fail to correct
the simulations in spite of the switch prediction. Once the first switch
occurs the simulation will change behavior and hence the first red cross marks
the point at which the simulation can be rejected from being correct.

After having made these observations we furthermore tested out a third way of
creating the scheme by predicting how long a timestep we could take before the
switch would occur and then divide this steplength by 2 to get a new timestep
length. This generated infeasibly small timesteps forcing us to stop the
simulation too early.

Taking these observations into account we can safely say that the
adapt-on-switch method/scheme is not working for our simulation. This does
however not mean that the method couldn't work for other domains or other
refinements of the method. The problem with the scheme is that the prediction
of switches is done too late in the simulation and hence once we detect
the switches it's too late to do anything about them given our model of
the mass-spring system. In order for the scheme to work properly it should
therefore be modified to be able to perform the detection of switches earlier
in the simulation.

In the real world the spring force would ``bend'' towards $\pm \infty$ once
the spring is stretched or contracted too much and hence the linear model that
we use doesn't suffice when using the adapt-on-switch scheme. Having used a
more realistic choice of model could possibly have had a positive impact of
the use of this scheme since this would cause the spring force of the spring
in question to grow in order to avoid the switch. \todo{Maybe the section
above should be moved to a discussion section.}

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/switch_multiscale_1fps_relative.png}
        \caption{Relative timestepping ($\Delta t = \frac{\Delta t}{100}$)}
        \label{fig:switch_multi_1fps_relative}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/switch_multiscale_1fps_absolute.png}
        \caption{Absolute timestepping ($\Delta t = \frac{1}{30}$)}
        \label{fig:switch_multi_1fps_absolute}
    \end{subfigure}
    \caption{Adapt-on-switch simulation of particles distributed with two
scales using 1 fps. The red crosses mark times at which the scheme predicts a
switch in the following iteration.}
    \label{fig:switch}
\end{figure}

\subsection{Atomic Updating using Heuristics}
The AUH method concists of two heuristics for determining whether an element
should be updated or not: steps since last update $l_{last}$, and the ratio
between the spring length in last update and the current length $r_{last}$.

We first test the method using only the first heuristic which will give
us synchronized updates of all vertices every $r_{last}$ step. Using the
same setup as the standard simulation we perform simulations using the AUH
method with the length ratio heuristic ignored by setting the limit of
the ratio to infinity. We set the timestep to $\Delta t = \frac{1}{10}$
(10 fps) and vary $l_{last}$. Figure \ref{fig:atomic_multi_sbound} shows
the resulting simulations where figures \ref{fig:atomic_multi_10sbound},
\ref{fig:atomic_multi_68sbound}, and \ref{fig:atomic_multi_69sbound} uses
an $l_{last}$ of 10, 68, and 69 respectively.

Simulating using 10 fps and $l_{last} = 10$ gives us the same amount
of velocity updates as the standard simulation using 1 fps (which had
switches), but gives us a nice and smooth simulation with no problems.
As can be seen in the figures \ref{fig:atomic_multi_68sbound} and
\ref{fig:atomic_multi_69sbound} we are able to perform the simulation using
$l_{last} = 68$ without having switches occuring whereas using $l_{last} =
69$ results in switches occuring. In other words we are able to perform the
simulation at 10 fps but with 6.8 seconds between acceleration and velocity
calculations. This is a result of combining the explicit Euler method with
the AUH scheme since the scheme simply extrapolates the position at the last
acceleration and velocity update to a position close in time to the next
update position and then use this position when updating the acceleration and
velocity again. Had we chosen an alternative time integration method this
advantage would most likely have been eliminated and let us perform the larger
timesteps with the standard simulation.

We now turn our attention towards the second heuristic.
\todo{Decide on the exact heuristic to display after having talked to Sune.}

\begin{figure}[H]
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=\textwidth]{../images/atomic_multiscale_10fps_10sbound.png}
        \caption{$l_{last} = 10$}
        \label{fig:atomic_multi_10sbound}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=\textwidth]{../images/atomic_multiscale_10fps_68sbound.png}
        \caption{$l_{last} = 68$}
        \label{fig:atomic_multi_68sbound}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=\textwidth]{../images/atomic_multiscale_10fps_69sbound.png}
        \caption{$l_{last} = 69$}
        \label{fig:atomic_multi_69sbound}
    \end{subfigure}
    \caption{Simulations using the AUH method at 10 fps with various number of
    steps since last update. Length heuristic is ignored (set to $\infty$).}
    \label{fig:atomic_multi_sbound}
\end{figure}

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/atomic_uniform_10fps_0.png}
        \caption{}
        \label{fig:atomic_uniform_10fps_0}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/atomic_uniform_10fps_1.png}
        \caption{}
        \label{fig:atomic_uniform_10fps_1}
    \end{subfigure}
\end{figure}

if ratio then too much dampening caused by lack of updates: Example: Simple with
3 springs

@@

The heuristic dampens the system
\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/inverse_uniform_10fps_0.png}
        \caption{}
        \label{fig:inverse_uniform_10fps_0}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/inverse_uniform_10fps_1.png}
        \caption{}
        \label{fig:inverse_uniform_10fps_1}
    \end{subfigure}
\end{figure}
